Performance analysis of high-resolution \\ ice sheet simulations

### 2022-10-08 paper/perfmod.tex
these edits complete my revised version, but I need to finish the responses document

In other words, these current-generation schemes implement \emph{explicit} time-stepping for the coupled mass and momentum system describing the dynamical evolution of ice sheets.\footnote{Confusingly, various ``semi-implicit'' and even ``fully implicit'' designators appear in the literature for explicit schemes which use velocity computed from the geometry at the start of the time step \citep[for example]{Chengetal2017}.}  However, at least for simpler, textbook partial differential equation problems, the limited and conditional stability of such explicit time-stepping schemes is well-understood \citep{LeVeque2007}.  Stability conditions of explicit SIA models appeared some time ago \citep[e.g.][]{HindmarshPayne1996}, but recent studies have focussed on the stability conditions of explicit hybrid, higher-order, and Stokes dynamics models \citep{Chengetal2017,Robinsonetal2022}, or on lengthening their steps \citep{LofgrenAhlkronaHelanow2021}.

However, actual \emph{implicit} time-stepping \citep{LeVeque2007} should also be considered.  Here the velocity and geometry are updated simultaneously by solving coupled mass and momentum conservation equations.  While such implicit time-stepping requires the solution of systems of equations at each step, for glaciers and ice sheets an implicit step must also \emph{simultaneously} compute the velocity field and the \emph{domain on which the velocity is defined}, namely the 3D extent of the ice once the coupled solution has converged.  The problem is of free-boundary type, especially in map-plane (horizontal) directions \citep{SchoofHewitt2013}, but also in the (easily-resolved) vertical direction.

An implicit strategy has been demonstrated at high (900m) resolution for the Greenlandic ice sheet in the simplest frozen-base, isothermal SIA case \citep{Bueler2016,JouvetGraeser2013}.  That work also shows how the steady-state SIA problem \citep[c.f.][]{JouvetBueler2012} can be solved, demonstrating unconditional stability.  (Observe that steady-state equations correspond to an implicit time step of infinite duration.)

@@

This is the context in which the current note relates time-stepping and stress-balance choices to computational effort.  The simplified performance analysis here exposes the most important considerations and trade-offs.  While this author expects that implicit time-stepping ice sheet models will eventually be the fastest, and that advanced solver techniques like multigrid \citep{Briggsetal2000} will be necessary, such beliefs should be assessed quantitatively to the extent possible.

@@

Thus, as can be addressed by linearized analysis \citep{Robinsonetal2022}, when thickness perturbations grow unstably under explicit time-stepping, i.e.~with too large a step, they do so by a mix of (discretized) advective and diffusive mechanisms.  Let $s(t,\bx)=H(t,\bx)+b(\bx)$ denote the surface elevation, for bed elevation $b(\bx)$.  A numerical thickness perturbation will often cause the velocity $\bu$ to respond by increasing in a direction close to downhill ($-\gradx s$), a direction correlated to $-\gradx H$ over large areas of an ice sheet.  In membrane-stress-resolving models like Stokes this happens through the non-local solution of the stress balance, in which the gravitational source term effectively acts along the surface gradient $\gradx s$ because of the ice-domain geometry.  One might write that Equation \eqref{eq:masscontinuity} has velocity $\bu(H,\gradx s)$, a non-local function of geometry; a stress balance solution is required to evaluate this function.  A numerical instability of Equation \eqref{eq:masscontinuity} occurs when the ice thickness under/over-shoots its correct value because the numerical velocity from evaluating this non-local function is too strong for the given time step.

@@

However, either statement \eqref{eq:masscontinuity} or \eqref{eq:siamasscontinuity} of the mass continuity equation is fundamentally incomplete without modeling the changing (map-plane) boundary position of the glacier or ice sheet.  The problem of determining the updated margin position is only well-posed via an inequality constraint, namely that ice thickness is nonnegative ($H\ge 0$), equivalently that the ice surface elevation equals or exceeds the bed elevation ($s \ge b$).  The mathematical role of this constraint has been understood for some time in the SIA model \citep{Calvoetal2002,JouvetBueler2012,SchoofHewitt2013}, but its role as the source of the margin location is universal across fluid-layer problems with signed source terms \citep{Bueler2021conservation}.

It is in this context that we can distinguish time-stepping types in coupled geometry-velocity, i.e.~mass and momentum conserving, glacier and ice sheet models.  A coupled model computes (\emph{fully}) \emph{implicit} time steps if no significant aspects of the geometry or velocity are held fixed during the (coupled) solution.  That is, in an implicit scheme a coupled and well-posed model for the updated ice surface elevation (or thickness) and velocity fields is solved.  A scheme which does not allow the map-plane ice-covered region to change during a time step cannot respond physically (conservatively) to a change in climate inputs; such a scheme is not actually implicit for the coupled problem and it cannot validly compute long-duration time steps.  Thus we say that a coupled scheme is \emph{semi-implicit} if, in particular, aspects of the ice geometry are held fixed during its coupled geometry-velocity solution, or if the velocity is held fixed during a time step as the geometry is updated.  A scheme which computes margin advance or retreat after the velocity field is updated is only semi-implicit, for example.  Finally, a scheme is \emph{explicit} for the coupled problem, and uses a time-step \emph{splitting} scheme \citep{HundsdorferVerwer2003}, if the ice geometry is held fixed during the velocity solution, and then this (accepted) velocity solution is used to update the geometry, e.g.~through an advection step for Equation \eqref{eq:masscontinuity}.

Thus a key point about time-stepping schemes for the coupled geometry and velocity problem for glaciers and ice sheets is that implicit schemes are not merely solving fixed-domain systems of coupled PDEs.  The mathematical problem for the coupled and implicit time-step includes the condition which controls the free boundary, namely the inequality constraint of nonnegative thickness \citep{Bueler2021conservation}.  Implicit schemes which solve these free-boundary problems at each time step can, at high spatial resolution, transcend stability limitations on time step duration.  They can do time-stepping at a rate controlled by climatic-input time scales.

@@

Table \ref{tab:notation} lists the parameters used in the numerical model performance analysis of this section.  The primary parameters are $\Delta x$, which is a representative value for the horizontal mesh (grid) cell diameter, and $m$, which is the number of nodes (vertices) in the horizontal mesh.  \emph{High resolution} refers to the equivalent $\Delta x\to 0$ and $m\to\infty$ limits; asymptotic and big-O notation is used only in this limit.

\begin{table*}[ht]
{\normalsize
\begin{tabular}{cll}
\emph{name} & \emph{meaning} & \emph{units} \\ \hline
$\alpha$    & one fixed-geometry Stokes velocity solution is $O(n^{1+\alpha})${\large \strut} work\\
$\beta$     & one implicit SIA geometry (and velocity) solution is $O(n^{1+\beta})$ work \\
$\gamma$    & one implicit, coupled Stokes geometry and velocity solution is $O(n^{1+\gamma})$ work \\
$D$         & representative geometric (SIA) diffusivity of an ice sheet & $\text{km}^2 \text{a}^{-1}$ \\
$L$         & width of map-plane domain & km \\
$m$         & degrees of freedom: number of nodes in the horizontal mesh \\
$q$         & time steps per model year needed to resolve modeled climate interactions & $\text{a}^{-1}$ \\
$\Delta t$  & length of time step in model years & a \\
$U$         & representative horizontal ice velocity & $\text{km}\,\text{a}^{-1}$ \\
$\Delta x$  & representative width (diameter) of map-plane mesh cells & km
\end{tabular}
}
\caption{Parameters for performance analysis.  Note $\alpha$, $\beta$, $\gamma$, and $m$ are pure, unit-less numbers.}
\label{tab:notation}
\end{table*}

Assuming the map-plane extent of the model domain is 2D and of width $L$, these primary parameters are related by
\begin{equation}
\Delta x \sim \frac{L}{m^{1/2}} \quad \text{and} \quad m \sim \frac{L^2}{\Delta x^2}.  \label{eq:paramrelation}
\end{equation}
That is, for fixed domain width $L$ there are $O(m^{1/2})$ mesh cells in each horizontal dimension.\footnote{In flow-line models $\Delta x = O(L m^{-1})$, but this paper addresses 3D models with map-plane (2D) horizontal meshes.}

@@

Such models also have $O(m)$ velocity variables, but note that these are not state variables.  That is, a very-viscous stress balance computes velocity as a function of the true state variables, especially ice thickness or surface elevation.

The above assumption of fixed vertical resolution reflects common usage \citep[for example]{Aschwandenetal2019,BrinkerhoffJohnson2015,Hoffmanetal2018,Lengetal2012,
Winkelmannetal2011}, and it permits a rational comparison of asymptotics, but it is not the only possibility.  Some solvers use 3D refinement \citep{BrownSmithAhmadia2013,IsaacStadlerGhattas2015,Tuminaroetal2016}, with various distinctions between how horizontal and vertical meshing is handled.  This and many other details cannot be pursued here.

Glacier and ice sheet models necessarily resolve and integrate ice-climate interactions, especially via surface mass balance, on time scales which are dominated by an annual cycle, and on longer scales.  Let $q$ be the number of ice-dynamical time steps per year needed to capture this coupling.  Typical values $q=0.1 \,\text{a}^{-1}, 1 \,\text{a}^{-1}, 12 \,\text{a}^{-1}$ correspond to decadal, yearly, and monthly frequency, respectively.  Note that energy balance and degree-day schemes for computing surface mass balance generally have much shorter time scales, but here $q$ describes the \emph{coupling} frequency on which ice geometry needs to be updated via solution of the mass continuity or surface kinematical equation.

@@

For an explicit SIA model the well-known stability restriction is $\Delta t < O(D^{-1} \Delta x^2)$ \citep{Bueleretal2005,HindmarshPayne1996}, where $D$ is a representative diffusivity value, i.e.~of $d$ in equation \eqref{eq:siamasscontinuity}.  For Stokes or other membrane-stress-resolving dynamics the stability of explicit time-stepping is largely unexplored in any precise sense, but an advective restriction $\Delta t < O(U^{-1} \Delta x)$, for some representative horizontal velocity scale $U$, might be said to represent the \emph{optimistic} paradigm.  The corresponding \emph{pessimistic} paradigm requires $\Delta t < O(D^{-1} \Delta x^2)$, using a representative diffusivity value $D$ computed in the SIA manner.

Explicit time-stepping with hybrid and higher-order schemes is somewhat better-studied than for Stokes dynamics, especially over horizontal resolutions relevant to whole ice sheets.  Some hybrid schemes apply the pessimistic paradigm as an adaptive restriction \citep{Winkelmannetal2011} across all spatial scales.  Other models apparently require the user to choose a fixed time step through trial and error in some circumstances \citep[for example]{Fischleretal2022,Robinsonetal2022}.  The optimistic paradigm has theoretical support for a certain higher-order DIVA scheme \citep[Equation (52)]{Robinsonetal2022}, but practical Greenland simulations in the same work actually suggest an intermediate power $\Delta t \approx O(\Delta x^{1.6})$ \citep[Figure 3(a)]{Robinsonetal2022}.  An intermediate power, a restriction $\Delta t = O(\Delta x^\omega)$ for some fuzzy value $1.5<\alpha<2$, also aligns with this author's experience.

Unconditionally-stable implicit schemes also have a maximum time step restriction, namely $\Delta t < O(q^{-1})$, but this restriction reflects the simulation purpose, not maintenance of stability.  For an implicit scheme the desired frequency of ice-climate interaction determines the total simulation cost only according to the (large) solution cost, at each time step, of solving coupled mass and momentum equations, including the surface kinematical or mass continuity equation.

@@

For each explicit time step of a model which applies a membrane-stress-resolving balance, the computational cost of a velocity (or velocity/pressure) solution of the stress balance equations is determined by solver design.  Let us assume that one such solution requires $O(m^{1+\alpha})$ floating point operations (flops),  with the power $\alpha\ge 0$ depending on the solver implementation.  For example, a Stokes solver using direct linear algebra for each Newton step might yield $\alpha \approx 1$ if sparsity is exploited or $\alpha \approx 2$ if not \citep{Bueler2021}.  However, a multigrid method \citep{Trottenbergetal2001} can greatly reduce $\alpha$, and the ideal value $\alpha=0$ describes an \emph{optimal} solver in the language of algorithmic scaling or solver complexity \citep{Bueler2021}.  For example, Antarctic ice sheet results by \cite{IsaacStadlerGhattas2015}, for a Stokes solver implemented using algebraic multigrid, show that the total number of preconditioned Krylov iterations \citep{Bueler2021}, over the nonlinear solve, grows slowly under mesh refinement, suggesting perhaps $\alpha\approx 0.2$ \citep[Table 8.1]{IsaacStadlerGhattas2015}.  The \cite{Lengetal2012} algebraic-multigrid Stokes solver may have similar scaling, but reported results do not constrain $\alpha$.  For a higher-order stress balance on the Greenlandic ice sheet, \cite{Tuminaroetal2016} report total algebraic-multigrid-preconditioned iterations suggesting $\alpha \approx 0.05$,\footnote{See \citep[Table 7.5]{Tuminaroetal2016}.  Somewhat worse performance for the Antarctic ice sheet is diagnosed as caused by the difficulties in discretizing a marine margin.} and a geometric multigrid method by \citep{BrownSmithAhmadia2013} suggests $\alpha$ is close to zero for simplified geometries.  Observe that the SIA velocity computation, a trivialization of the Stokes problem in which velocity is computed by a pointwise formula, requires optimal $O(m)$ flops.

When analyzing solver scaling in our simplified form, one must be aware that the constant in $O(m^{1+\alpha})$ can be very large, and it depends strongly on solver design.  Furthermore, many considerations are suppressed in any flops-based analysis of algorithmic scaling.  Actual run time is also determined by memory latency, memory bandwidth, and process/thread/GPU parallelism, among other factors.  Also, the \cite{BrownSmithAhmadia2013,Fischleretal2022,IsaacStadlerGhattas2015,Lengetal2012,Tuminaroetal2016} results, among others, all show good \emph{parallel} scaling, something not addressed here.

Regardless of the stress balance, an explicit time-stepping scheme will first compute velocity from current geometry and then apply the mass continuity equation to update the ice thickness using $O(m)$ work.  That is, once the velocity is computed from the last-known geometry, an explicit scheme replaces old thickness values by new ones using a pointwise formula.\footnote{Any additional computation needed to remesh the updated geometry, a highly design-dependent cost, is omitted here.}

@@

Now, how many flops are needed to simulate one model year?  Suppose a numerical model takes time steps of $\Delta t$ model years duration, equivalently $\Delta t^{-1}$ steps per model year.  In these terms one may write stability restrictions as required numbers of steps per model year.  That is, stability requires that $\Delta t^{-1}$ be bounded below by a function of the horizontal resolution $\Delta x$ or the degrees of freedom $m$.  Recalling the scaling in Equation \eqref{eq:paramrelation}, of $\Delta x$ with $m$, the explicit SIA and pessimistic-Stokes cases have
\begin{equation}
\frac{1}{\Delta t} > \oo{\frac{D}{\Delta x^2}} = \oo{\frac{D m}{L^2}}. \label{eq:explicitsiarequired}
\end{equation}
The explicit, optimistic-Stokes estimate becomes
\begin{equation}
\frac{1}{\Delta t} > \oo{\frac{U}{\Delta x}} = \oo{\frac{U m^{1/2}}{L}}. \label{eq:explicitoptstokesrequired}
\end{equation}

The number of time steps per model year is then multiplied by the per-step computational cost, namely $O(m^{1+\alpha})$, to give a work estimate for each model year in a simulation.  The results so far are shown in the ``explicit'' rows of Table \ref{tab:performancemodel}.

As already explained, unconditionally-stable implicit methods have a fixed time step $\Delta t = 1/q$, independent of $\Delta x$ and determined only by the need to resolve ice-climate interactions.  On the other hand, the per-step expense is much greater because nontrivial coupled equations, and a free-boundary problem, must be solved for the velocity and geometry-update simultaneously.  For SIA models, let us assume that the flops of such coupled solutions scale as $O(m^{1+\beta})$ with $\beta \ge 0$; a large constant is assumed.

% regarding next paragraph, see data/bueler.perf and data/scaling.m

The only currently implemented, unconditionally-stable, fully-implicit geometric-update solvers use the SIA stress balance.  For simplified (dome) geometry the scheme in \cite{Bueler2016}, based on Newton steps solved via direct linear algebra and single-grid ice margin determination, shows $\beta=0.8$.  (An earlier steady, Picard iteration implementation by \cite{JouvetBueler2012} scales worse.)  Convergence is also robust for implicit time steps of years to centuries on kilometer-scale grids for the Greenland ice sheet, using realistic, irregular, bed topographies.

>>>>

In other words, these current-generation schemes implement \emph{explicit} time-stepping for the coupled mass and momentum system describing the dynamical evolution of ice sheets.\footnote{Confusingly, various ``semi-implicit'' and even ``fully implicit'' designators appear in the literature for explicit schemes which fix the velocity following the computation which updates the geometry \citep[for example]{Chengetal2017}.}  However, at least for simpler, textbook partial differential equation problems, the limited and conditional stability of such explicit time-stepping schemes is well-understood \citep{LeVeque2007}.  Stability conditions of explicit SIA models appeared some time ago \citep[e.g.][]{HindmarshPayne1996}, but recent studies have focussed on the stability conditions of explicit hybrid, higher-order, and Stokes dynamics models \citep{Chengetal2017,Robinsonetal2022}, or on lengthening their steps \citep{LofgrenAhlkronaHelanow2021}.

However, actual \emph{implicit} time-stepping \citep{LeVeque2007} should also be considered.  Here the velocity and geometry are updated simultaneously by solving coupled mass and momentum conservation equations.  While such implicit time-stepping requires the solution of systems of equations at each step, for glaciers and ice sheets an implicit step must simultaneously compute the velocity field and the \emph{domain on which the velocity is defined}, namely the 3D extent of the ice once the coupled solution has converged.  The problem is of free-boundary type, especially in map-plane (horizontal) directions \citep{SchoofHewitt2013}, but also in the (easily-resolved) vertical direction.

An implicit strategy has been demonstrated at high (900m) resolution for the Greenlandic ice sheet in the simplest frozen-base, isothermal SIA case \citep{Bueler2016}.  That work also shows how the steady-state SIA problem \citep[c.f.][]{JouvetBueler2012} can be solved, demonstrating unconditional stability.  (Observe that steady-state equations correspond to an implicit time step of infinite duration.)

@@

This is the context in which the current note relates time-stepping and stress-balance choices to computational effort.  The simplified performance analysis here exposes the most important considerations and trade-offs.  While this author expects that implicit time-stepping ice sheet models will eventually be the fastest, and that advanced solver techniques like multigrid \citep{Briggsetal2000} will lead to better modeling, such beliefs should be assessed quantitatively to the extent possible.

@@

Thus, as can be addressed by linearized analysis \citep{Robinsonetal2022}, when thickness perturbations grow unstably under explicit time-stepping, i.e.~with too large a step, they do so by a mix of  advective and diffusive mechanisms.  Let $s(t,\bx)=H(t,\bx)+b(\bx)$ denote the surface elevation, for bed elevation $b(\bx)$.  A numerical thickness perturbation will often cause the velocity $\bu$ to respond by increasing in a direction close to downhill ($-\gradx s$), a direction correlated to $-\gradx H$ over large areas of an ice sheet.  In membrane-stress-resolving models like Stokes this happens through the non-local solution of the stress balance, in which the gravitational source term effectively acts along the surface gradient $\gradx s$ because of the ice-domain geometry.  One might write that Equation \eqref{eq:masscontinuity} has velocity $\bu(H,\gradx s)$, a non-local function of geometry; a stress balance solution is required to evaluate this function.  A numerical instability of Equation \eqref{eq:masscontinuity} occurs when the ice thickness under/over-shoots its correct value because the numerical velocity from evaluating this non-local function is too strong for the given time step.

@@

However, either statement \eqref{eq:masscontinuity} or \eqref{eq:siamasscontinuity} of the mass continuity equation is fundamentally incomplete without modeling the changing (map-plane) boundary position of the glacier or ice sheet.  The problem of determining this updated margin position is only well-posed via an inequality constraint, namely that ice thickness is nonnegative ($H\ge 0$), equivalently that the ice surface elevation equals or exceeds the bed elevation ($s \ge b$).  The mathematical role of this constraint has been understood for some time in the SIA model \citep{Calvoetal2002,JouvetBueler2012,SchoofHewitt2013}, but its role as the determinant of margin location is universal across fluid-layer problems with signed source terms \citep{Bueler2021conservation}.

In this context one can distinguish time-stepping types in coupled geometry-velocity, i.e.~mass and momentum conserving, glacier and ice sheet models.  A coupled model computes (\emph{fully}) \emph{implicit} time steps if no significant aspects of the geometry or velocity are held fixed during the (coupled) solution.  That is, in an implicit scheme a coupled and well-posed model for the ice surface elevation (or thickness) and velocity field updates is solved.  A scheme which does not allow the map-plane ice-covered region to change during the time step cannot respond physically (conservatively with respect to mass and momentum conservation) to a change in climate inputs.  A scheme which is implicit in the sense here can be unconditionally stable; it can stably compute time steps of arbitrary duration causing nontrivial changes in margin position.  Next, a coupled scheme is \emph{semi-implicit} if, in particular, aspects of the ice geometry are held fixed during the geometry-velocity solution, for instance if the velocity is held fixed during a time step as the geometry is updated (perhaps ``implicitly'' on its own).  Thus a scheme which computes margin advance or retreat only after the velocity field update is accepted is only semi-implicit, for example.  Finally, a scheme is \emph{explicit} for the coupled problem if the ice geometry is held fixed during the velocity solution, and then this (accepted and now fixed) velocity solution is used to update the geometry through an explicit step for Equation \eqref{eq:masscontinuity}.

Thus a key point about time-stepping schemes for the coupled geometry and velocity problem for glaciers and ice sheets is that implicit schemes are not solving fixed-domain systems of coupled PDEs.  The mathematical problem for the coupled and implicit time-step includes the condition which controls the moving and free boundary, namely the inequality constraint of nonnegative thickness.  Implicit schemes which solve these free-boundary problems at each time step can, at high spatial resolution, transcend stability limitations on time step duration.  They, and only they, can do time-stepping at a rate controlled only by ice-climate interaction time scales.

@@

Table \ref{tab:notation} lists the parameters used in the numerical model performance analysis which follows.  The primary parameters are $\Delta x$, which is a representative value for the horizontal mesh (grid) cell diameter, and $m$, the number of nodes (vertices) in the horizontal mesh.  \emph{High resolution} refers to the equivalent $\Delta x\to 0$ and $m\to\infty$ limits; asymptotic and big-O notation is used only in this limit.

\begin{table*}[ht]
{\normalsize
\begin{tabular}{cll}
\emph{name} & \emph{meaning} & \emph{units} \\ \hline
$\alpha$    & one fixed-geometry Stokes velocity solution is $O(n^{1+\alpha})${\large \strut} work\\
$\beta$     & one implicit SIA geometry (and velocity) solution is $O(n^{1+\beta})$ work \\
$\gamma$    & one implicit, coupled Stokes geometry and velocity solution is $O(n^{1+\gamma})$ work \\
$D$         & representative geometric (SIA) diffusivity of an ice sheet & $\text{km}^2 \text{a}^{-1}$ \\
$L$         & width of map-plane domain & km \\
$m$         & degrees of freedom: number of nodes in the horizontal mesh \\
$q$         & time steps per model year needed to resolve ice-climate interactions & $\text{a}^{-1}$ \\
$\Delta t$  & length of time step & a \\
$U$         & representative horizontal ice velocity & $\text{km}\,\text{a}^{-1}$ \\
$\Delta x$  & representative width (diameter) of map-plane mesh cells & km
\end{tabular}
}
\caption{Parameters for performance analysis.  Note $\alpha$, $\beta$, $\gamma$, and $m$ are pure, unit-less numbers.}
\label{tab:notation}
\end{table*}

Assuming the map-plane extent of the model domain is 2D and of width $L$, these primary parameters are related by
\begin{equation}
\Delta x \sim \frac{L}{m^{1/2}} \quad \text{and} \quad m \sim \frac{L^2}{\Delta x^2}.  \label{eq:paramrelation}
\end{equation}
Specifically, for fixed domain width $L$ there are $O(m^{1/2})$ mesh cells in each horizontal dimension.\footnote{In flow-line models $\Delta x = O(L m^{-1})$, but this paper addresses 3D models with map-plane (2D) horizontal meshes.}

@@

Such models also have $O(m)$ velocity variables, but note that these are not state variables.  That is, a very-viscous stress balance computes velocity as a function of the true state variables, such as ice thickness and temperature.

The above assumption of fixed vertical resolution reflects common usage \citep[for example]{Aschwandenetal2019,BrinkerhoffJohnson2015,Hoffmanetal2018,Lengetal2012,
Winkelmannetal2011}, and it permits a rational comparison of asymptotics, but it is not the only possibility.  Some solvers use 3D refinement \citep{BrownSmithAhmadia2013,IsaacStadlerGhattas2015,Tuminaroetal2016}, with various distinctions between how horizontal and vertical meshing is handled, but this and many other details cannot be pursued here.

Glacier and ice sheet models resolve and integrate ice-climate interactions, especially via surface mass balance, on time scales which are dominated by an annual cycle, and on longer scales.  Let $q$ be the number of ice-dynamical time steps per year needed to capture this coupling.  Typical values $q=0.1 \,\text{a}^{-1}, 1 \,\text{a}^{-1}, 12 \,\text{a}^{-1}$ correspond to decadal, yearly, and monthly frequency, respectively.  Note that energy balance and degree-day schemes for computing surface mass balance \citep{GreveBlatter2009} generally have much shorter time scales, but here $q$ describes the \emph{coupling} frequency on which ice geometry needs to be updated via solution of the mass continuity or surface kinematical equation.

@@

For an explicit SIA model the well-known stability restriction is $\Delta t < O(D^{-1} \Delta x^2)$ \citep{Bueleretal2005,HindmarshPayne1996}, where $D$ is a representative diffusivity value, i.e.~of $d$ in equation \eqref{eq:siamasscontinuity}.  For Stokes models, or other membrane-stress-resolving dynamics, the stability of explicit time-stepping is largely unexplored in any precise sense, but an advective restriction $\Delta t < O(U^{-1} \Delta x)$, for some representative horizontal velocity scale $U$, might be said to represent the \emph{optimistic} paradigm.  The corresponding \emph{pessimistic} paradigm for Stokes models says $\Delta t < O(D^{-1} \Delta x^2)$, using a representative diffusivity value $D$ computed in the SIA manner.

Explicit time-stepping with hybrid and higher-order schemes is somewhat better-studied than for Stokes dynamics, especially over horizontal resolutions relevant to whole ice sheets.  Some hybrid schemes apply the pessimistic paradigm as an adaptive restriction \citep{Winkelmannetal2011} across all spatial scales.  Other models apparently require the user to choose a fixed time step through trial and error in some circumstances \citep[for example]{Fischleretal2022,Robinsonetal2022}.  The optimistic paradigm has theoretical support for a certain higher-order DIVA scheme \citep[Equation (52)]{Robinsonetal2022}, but practical Greenland simulations in the same work actually suggest an intermediate power $\Delta t \approx O(\Delta x^{1.6})$ \citep[Figure 3(a)]{Robinsonetal2022}.  An intermediate power, a restriction $\Delta t = O(\Delta x^\omega)$ for some fuzzy value $1.5<\omega<2$, also aligns with this author's experience.

Unconditionally-stable implicit schemes also have a maximum time step restriction, namely $\Delta t < O(q^{-1})$, but this restriction reflects the simulation purpose, not maintenance of stability.  For an implicit scheme the desired frequency of ice-climate interaction will, however, determine the total simulation cost according to the (large) solution cost, at each time step, of solving coupled mass and momentum equations, including the surface kinematical or mass continuity equation.

@@

For each explicit time step of a model which applies a membrane-stress-resolving balance, the computational cost of a velocity (or velocity/pressure) solution of the stress balance equations is determined by solver design.  Let us assume that one such solution requires $O(m^{1+\alpha})$ floating point operations (flops),  with the power $\alpha\ge 0$ depending on the solver implementation.  For example, a Stokes solver using direct linear algebra for each Newton step might yield $\alpha \approx 1$ if sparsity is exploited or $\alpha \approx 2$ if not \citep{Bueler2021}.  However, a multigrid method \citep{Trottenbergetal2001} can greatly reduce $\alpha$, and the ideal value $\alpha=0$ describes an \emph{optimal} solver in the language of algorithmic scaling or solver complexity \citep{Bueler2021}.  For example, Antarctic ice sheet results by \cite{IsaacStadlerGhattas2015}, for a Stokes solver implemented using algebraic multigrid, show that the total number of preconditioned Krylov iterations, over the nonlinear solve, grows slowly under mesh refinement, suggesting perhaps $\alpha\approx 0.2$ \citep[Table 8.1]{IsaacStadlerGhattas2015}.  The \cite{Lengetal2012} algebraic-multigrid Stokes solver may have similar scaling, but reported results do not constrain $\alpha$.  For a higher-order stress balance on the Greenlandic ice sheet, \cite{Tuminaroetal2016} report total algebraic-multigrid-preconditioned iterations suggesting $\alpha \approx 0.05$,\footnote{See \citep[Table 7.5]{Tuminaroetal2016}.  Somewhat worse performance for the Antarctic ice sheet is diagnosed as caused by the difficulties in discretizing a marine margin.} and a geometric multigrid method by \citep{BrownSmithAhmadia2013} suggests $\alpha$ is close to zero for simplified geometries.  Observe that the SIA velocity computation, a trivialization of the Stokes problem in which velocity is computed by a pointwise formula, requires optimal $O(m)$ flops.

When analyzing solver scaling in our simplified form, one must be aware that the constant in $O(m^{1+\alpha})$ can be very large, and it depends strongly on solver design.  Furthermore, many considerations are suppressed in any flops-based analysis of algorithmic scaling, as actual run time is also determined by memory latency, memory bandwidth, and process/thread/GPU parallelism, among other factors.  Many models \citep[e.g.][]{BrownSmithAhmadia2013,Fischleretal2022,IsaacStadlerGhattas2015,Lengetal2012,Tuminaroetal2016} also show good \emph{parallel} scaling, something not addressed here.

Regardless of the stress balance, an explicit time-stepping scheme will first compute velocity from current geometry and then apply the mass continuity equation to update the ice thickness using $O(m)$ work.  That is, once the velocity is computed from the last-known geometry, let us assume that an explicit scheme replaces old thickness values by new ones using an optimal pointwise formula.\footnote{Any additional computation needed to remesh the updated geometry, a highly design-dependent cost, is omitted here.}

@@

Now, how many flops are needed to simulate one model year?  Suppose a numerical model takes time steps of $\Delta t$ model years duration, equivalently $\Delta t^{-1}$ steps per model year.  In these terms stability restrictions give required numbers of steps per model year.  That is, stability requires that $\Delta t^{-1}$ be bounded below by a function of the horizontal resolution $\Delta x$ or the degrees of freedom $m$.  Recalling the scaling in Equation \eqref{eq:paramrelation}, the explicit SIA and pessimistic-Stokes cases have
\begin{equation}
\frac{1}{\Delta t} > \oo{\frac{D}{\Delta x^2}} = \oo{\frac{D m}{L^2}}. \label{eq:explicitsiarequired}
\end{equation}
The explicit, optimistic-Stokes estimate becomes
\begin{equation}
\frac{1}{\Delta t} > \oo{\frac{U}{\Delta x}} = \oo{\frac{U m^{1/2}}{L}}. \label{eq:explicitoptstokesrequired}
\end{equation}
The number of time steps per model year is then multiplied by the per-step computational cost, namely $O(m^{1+\alpha})$ for Stokes models, to give a work estimate for each model year in a simulation.  The results so far are shown in the ``explicit'' rows of Table \ref{tab:performancemodel}.

As already explained, unconditionally-stable implicit methods may have a fixed time step $\Delta t = 1/q$, independent of $\Delta x$ and determined only by the need to resolve ice-climate interactions.  On the other hand, the per-step expense is much greater because nontrivial coupled equations, and a free-boundary problem, must be solved simultaneously for the updated velocity and geometry.

% regarding next paragraph, see data/bueler.perf and data/scaling.m
For SIA models, let us assume that the flops of such coupled solutions scale as $O(m^{1+\beta})$ with $\beta \ge 0$; a large constant is assumed.  The only currently implemented, unconditionally-stable, fully-implicit geometric-update solvers use the SIA stress balance.  For simplified (dome) geometry the scheme in \cite{Bueler2016}, based on Newton steps solved via direct linear algebra and single-grid ice margin determination, shows $\beta=0.8$.  (An earlier steady, Picard iteration implementation by \cite{JouvetBueler2012} scales worse.)  Convergence is also robust for implicit time steps of years to centuries on kilometer-scale grids for the Greenland ice sheet, using realistic and irregular bed topographies.

Certain recent numerical models are semi-implicit in innovative ways.  The SIA portions of hybrid time-stepping schemes by \cite{JouvetGraeser2013} and \cite{BrinkerhoffJohnson2015} are solved implicitly, with nonnegativity of thickness in their SIA portions enforced as part of a free-boundary solution.  However, the time-stepping in these models is only semi-implicit because the membrane-stress-resolving portion of the velocity explicitly advects thickness.  The SIA time-step solvers in these schemes have apparently only been tested for time steps satisfying an advective condition $\Delta t < O(U^{-1}\Delta x)$, for $U$ derived from sliding speeds.  (Also, \cite{JouvetGraeser2013} use multigrid methods, but their published results do not constrain the power $\beta$.)
% equation (20) in JouvetGraeser2013
% equation (A21) in BrinkerhoffJohnson2015
In a different design, the semi-implicit scheme in the open-source \'Ua numerical model \citep{Gudmundsson2021manual} implicitly solves mass continuity Equation \eqref{eq:masscontinuity} using an active set method, but while holding the membrane-stress-resolved velocity fixed.  Note that the formally third-order time-stepping in this model \citep{Gudmundsson2013}, while surely desirable for marine ice sheets, cannot achieve that accuracy when the nonnegativity-of-thickness constraint is actively determining the free boundary, e.g.~at retreating margins.
% equations (1), (26) in Gudmundson2013
% see semi-implicit comments in commit 9081fdbe3cf250d9cfd66504bf03134043ba2547
% https://github.com/GHilmarG/UaSource/blob/master/NexthTG3in2D.m: "Third order impicit Taylor-Galerkin (implict with respect to h, not u and v)"
% https://github.com/GHilmarG/UaSource/blob/master/SSS2dPrognostic.m: "semi-implicit forward integration. implicit with respect to thickness (h) explicit with respect to velocity"
% and "error('Semi-implicit time integration only implemented for SSTREAM and not SSHEET or Hybrid formulations \n')"

For implicit Stokes time-stepping, a coupled and free-boundary velocity and geometry-update solve is assumed to be, in the absence of constraining research, $O(m^{1+\gamma})$ for some $\gamma \ge \alpha$ to be determined.  One might also suppose $\gamma\ge \beta$, but there are no implemented cases to measure.  These comments complete Table \ref{tab:performancemodel}.  Note that all implicit scheme estimates involve an especially-large scheme-dependent constant.
